---
title: "In-Class Programming Activity 13"
author: 'Math 253: Statistical Computing & Machine Learning'
date: "k-fold cross-validation"
params: 
  show_answers: FALSE
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    latex_engine: xelatex
  tufte::tufte_book:
    latex_engine: xelatex
---

---

```{r include=FALSE}
options(width=80)
library(tufte)
library(mosaicData)
library(ggplot2)
library(scoreActivity)
library(ISLR)
library(dtkbook)
knitr::opts_chunk$set(tidy=FALSE, echo=params$show_answers)
options(width=100)
```

Today you are going to write a function to carry out k-fold cross validation.
Recall that this involves dividing the available data set into k equal-sized, independent sets.  For each of the k sets, reserve that set as testing data, fit your model to the remaining data, and evaluate the fitted model on the reserved set.  In doing this, you will accumulate k different estimates of the model error.  Average those to produce an overall estimate of model error.

# A function framework

Call your function `k_fold1()`.  It should have this interface:
```{r echo=TRUE}
k_fold1 <- function(formula, method = lm, data = mtcars, predfun = predict, k = 10) {
  # do the calculations, 
  # producing an estimate of error
  return(error_estimate)
}
```

As written above, the function does nothing.  You're going to fix that.  But use exactly the interface (function name, argument names and order, default values) given above.

# The k sets

In k-fold cross validation, you'll divide the cases in your data into k sets. To help do that, you're going to create a vector named `sets` that has one element for each case in the data. The value of that element will be 1 if the corresponding case is to be in set 1 of the k sets, 2 if the case is to be in set 2, and so on.  It doesn't matter how you assign cases to sets, so long as there are k sets of roughly equal size.

There are many ways to construct the vector `sets`.  `r margin_note(string_to_code("k <- 10\nsets <- (1:51 %% k) + 1\n# or, alternatively, ...\nsets <- rep(1:k, each = 51/k,\n      \t\t\t       length.out = 51)"))` Here are two operations, `%%` and `rep(, , each=)` that might be the basis for two different ways of creating the vector Here are two operations that may give you a hint. The example assumes that there are 51 rows in the `data`.


```{r results="hide", echo=FALSE}
k <- 10
sets <- (1:51 %% k) + 1
# or, alternatively, ...
sets <- rep(1:k, each = 51/k, length.out = 51)
```

# The loop

Your function will perform the same operation k times.  Doing this will produce k numbers, the mean square prediction error for each of the k test data sets.  In writing a loop, you ... 


1. Set up the state. Here's a code fragment that handles (1), (2), (3) and (5).
2. Construct the outline of the loop
3. Set up the initial state, or "accumulator." Make sure that you understand what `numeric(k)` is doing.*
4. Fill in the operations to be carried out inside the loop. 
5. Tidy up the accumulator to produce the final result.
```r
## accumulator will be updated as you go through the loop
mspe <- numeric(k)
for (i in 1:k) {     
  mspe[i] <- Your statements go here
}
mean(mspe)
```
 

```{r eval=FALSE, echo=FALSE}
mspe <- numeric(k)
for (i in 1:k) {
  # Your statements will go here
}
mean(mspe)
```



# Inside the loop

As you can see, the statements inside the loop will be evaluated k times.  The object `i` can be used inside the loop to indicate which of these k passes through the loop is currently being performed.

Here's what to do inside the loop:

1. Compare `i` to `sets` to create a logical (that is, `TRUE`/`FALSE` vector) of the rows to be used for the **test** set. Create a new data frame, `For_Testing` that has just these rows.
2. Using logical negation (that is, `!`) to create another data frame, fill a data frame `For_Training` with the remaining rows.
3. Fit your model using `For_Training`.  For simplicity in developing your function, you can use a particular model appropriate for the `mtcars` data set: `mod <- lm(mpg ~ hp + wt + am)`.  Note that `mpg` is the response variable.  Later on, you'll replace this with a more general statement.
4. Evaluate the model on the `For_Testing` data. You can do this with `pred_vals <- predict(mod, newdata = For_Testing)`
5. Calculate the mean square prediction error (MSPE). This will be `mean((For_Testing[["mpg"]] - pred_vals)^2 )`
6. Save the MSPE in the appropriate slot of `mspe`.


```{r echo = params$show_answers}
k_fold1 <- function(formula, method = lm, data = mtcars, predfun = predict, k = 10) {
  mspe <- numeric(k)
  sets <- (1:nrow(data) %% k) + 1 
  for (i in 1:k) {
    test_cases <- i == sets
    For_Testing  <- data[test_cases,]
    For_Training <- data[ ! test_cases,]
    mod <- lm(mpg ~ hp + wt + am, data = For_Training)
    pred_vals <- predict(mod, newdata = For_Testing)
    mspe[i] <- mean((For_Testing[["mpg"]] - pred_vals)^2)
  }

  return(mean(mspe))
}
```

# Trying out your function

At this point, you should have a function `k_fold1()` that, when evaluated with the default settings for the arguments, returns a number.  That number should be very roughly similar in size to this one, which is called the "in-sample" error.
```{r}
mod <- lm(mpg ~ hp + wt + am, data = mtcars)
mean(resid(mod)^2)
```
In-sample prediction errors are biased to be lower than cross-validated prediction errors.

# Generalizing the function

When you are satisfied that your `k_fold1()` function is working, *copy* the code to create another function which we'll call `k_fold()`.  This is going to be the generalization of the function to other data sets and other model formulas.

Modify `k_fold()` in the following ways:

* Replace `mpg ~ hp + wt + am` with `formula`.  This will allow you to specify the formula on as the first argument.
* Replace `lm()` with `method()`.  This will allow your function to use modeling types other than `lm()`
* Replace `predict()` with `predfun()`
* Replace `[["mpg"]]` with `[[as.character(formula[[2]])]]`.^[Indexing formula objects enables extraction of parts of the formula.  For those with an interest in computer languages, a formula can be used as a parse tree, with the operator (e.g. `~` or `+`, etc.) at the top and the arguments underneath.]

```{r echo = params$show_answers}
k_fold <- function(formula, method = lm, data = mtcars, predfun = predict, k = 10) {
  mspe <- numeric(k)
  sets <- (1:nrow(data) %% k) + 1 
  for (i in 1:k) {
    test_cases <- i == sets
    For_Testing  <- data[test_cases,]
    For_Training <- data[ ! test_cases,]
    mod <- method(formula, data = For_Training)
    pred_vals <- predfun(mod, newdata = For_Testing)
    mspe[i] <- mean((For_Testing[[as.character(formula[[2]])]] - pred_vals)^2)
  }

  return(mean(mspe))
}
```

Test out your function using the same data and formula as before, that is:
```{r eval = params$show_answers, echo=TRUE}
k_fold(formula = mpg ~ hp + wt + am, data=mtcars)
```

You should get the same answer as with `k_fold1()`.

```{r echo=FALSE, eval=params$show_answers, results="asis"}
cat("# Test statments")
```

```{r echo=params$show_answers, eval=params$show_answers}
scoreActivity::score253(12)
```
