---
title: "In-Class Computing Project 3"
author: 'Math 253: Statistical Computing & Machine Learning'
date: 'Working with indices: Training vs testing data'
output:
  tufte::tufte_handout:
    latex_engine: xelatex
  tufte::tufte_book:
    latex_engine: xelatex
  tufte::tufte_html: default
params:
  show_answers: no
---

```{r include=FALSE}
set.seed(101)
library(ISLR)
```


In today's programming project, you are going to build some models of the attractiveness of colleges to accepted applicants.  You'll use some data from the `ISLR` package; the package that goes along with the book.  You may need to install this package.^[If so, you should do it with the RStudio "Packages" tab.  **Do not** put the installation command in your `.Rmd` file.]

## Task 1

Load the `ISLR` package into R.  Then, to gain access to the `College` data in `ISLR`, use this command:
```{r}
data(College, package = "ISLR")
```

## Task 2

Create a variable called `Yield` within the `College` data table.  The yield is defined by college admissions officers to be the number of students enrolled divided by the number of students accepted. (Applications are another matter altogether.)

```{r echo = params$show_answers}
College$Yield <- with(College, Accept / Enroll)
```

## Task 3

Divide `College` into two data frames, one for training and testing.

* Create an object `all_indices` with the integers 1, 2, 3, $\ldots, n$, where $n$ is the number of rows in `College`
* Create an object `train_indices` with 200 random indices between 1 and the number of cases in `College`.  Hint: `sample()`
* Create another object `test_indices` with all the remaining cases from `College`. Hint: `setdiff()`.
* Create a data frame object `Train_data` with the rows from `College` corresponding to `train_indices`.  Hint: `College[ , ]`
* Create a data frame object `Test_data` with the rows from `College` corresponding to `test_indices`.

```{r echo = params$show_answers}
all_indices <- 1:nrow(College)
train_indices <- sample(all_indices, size = 200)
test_indices <- setdiff(all_indices, train_indices)
Train_data <- College[train_indices,]
Test_data <- College[test_indices,]
```

## Task 4

Using `Train_data`, construct a model of `Yield` as a function of `Top10perc`, `Outstate` (tuition), and  `Expend`. Arrange things so that the name of the object holding the model is `Yield_mod1`.

```{r}
Yield_mod1 <- lm(Yield ~ Top10perc + Outstate + Expend,
                 data=Train_data)
```

## Task 5

* Create an object `Y_train` which holds just the `Yield` from the training data. Hint: `Train_data$Yield`

```{r echo = params$show_answers}
Y_train <- Train_data$Yield
```

* Create an object `fhat_train` which is the output of the model for the inputs in the training data. Hint: `predict(Yield_mod1, newdata = Train_data )`

```{r echo = params$show_answers}
fhat_train <- predict(Yield_mod1, newdata = Train_data)
```

* Create an object `MSE_train` that holds the mean square error for the training data. The value contained in this object will be a single number.

```{r echo = params$show_answers}
MSE_train <- mean((Y_train - fhat_train)^2) 
```

## Task 6

Repeat Task 5, but for the testing data.  Everywhere `Train` or `train` appears in step 5, use `Test` or `test` in this step.  You'll end up with an object called `MSE_test`.

```{r echo = params$show_answers}
Y_test <- Test_data$Yield
fhat_test <- predict(Yield_mod1, newdata = Test_data)
MSE_test <- mean((Y_test - fhat_test)^2) 
```

You might be interested to look at the ratio of `MSE_train` to `MSE_test`. This will be random, but should be close to 1.

```{r echo = params$show_answers, results = "hide"}
MSE_train / MSE_test
```

```{r echo = params$show_answers, eval = params$show_answers}
scoreActivity::score253(3)
```
