---
title: "In-Class Computing Task 14"
author: "Math 253: Statistical Computing & Machine Learning"
date: "Shrinkage and out-of-sample prediction error"
params:
  show_answers: FALSE
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    latex_engine: xelatex
  tufte::tufte_book:
    latex_engine: xelatex
---

```{r include=FALSE}
options(width=80)
library(mosaicData)
library(ggplot2)
library(scoreActivity)
library(ISLR)
knitr::opts_chunk$set(tidy=FALSE, echo=params$show_answers)
options(width=100)
```

In this activity, you're going to examine how a shrinkage approach can reduce *out-of-sample* prediction error.

Install the `glmnet` package.^[Remember, *installing* a package is different than loading it with `libaray()`. You need to install a package only once, and it will work in any document. So don't put the installation command in your your document: use the Packages tab.  Your document should contain only the commands to load packages that have been installed previously.]

```{r message = FALSE}
library(glmnet)
```

You're going to use the `glmnet()` function to fit ridge regression models.
The data used in the examples can be loaded with 
```{r echo=TRUE}
GenCont <- read.csv("http://tiny.cc/dcf/GenCont.csv")
```

This is a data table about the expression of 12 genes and their possible relationship to the expression of a phenotype. The variable `Phenotypes` will be the response in your model. The 12 genes will be the explanatory variables.

\enlargethispage{1.5in}


# Comparing OLS to ridge

Write a function with this interface:
```{r echo=TRUE}
compare_ols_ridge <- function(responses, predictors, lambda=1){
  # your statements will go here
}
```

Your `compare_ols_ridge()` should do the following:

1. Create a training data set by choosing about half the cases at random. You will need to make both the responses and the predictors.
2. Create a testing data set which will be the remaining cases. Again, you'll need both the responses and the predictors.
3. Fit an ordinary linear regression using `lm()` on the training data.
4. Fit a ridge regression for the specified `lambda` using `glmnet()`. Set `alpha=0` as an argument to get ridge regression. 
5. For each of the models in (3) and (4), calculate the *in-sample* predictions using `predict()` and then turn these into an in-sample mean square prediction error.
6. Using the testing data set, calculate the *out-of-sample* mean square prediction error.
7. Your function should return a named vector with these five components: `lambda` as well as `ols_in`, `ridge_in`, `ols_out`, `ridge_out`, each of which has the corresponding mean square prediction error. 
8. Use your function to create an object called `Shrink_results` with `lambda = 1` and the `GenCont` data . 

```{r echo = params$show_answers}
compare_ols_ridge <- function(responses = GenCont[,1], 
                              predictors = GenCont[,-1], 
                              lambda=1) {
  training_inds <- sample(1:length(responses), size = floor(length(responses)/2))
  resp_train <- responses[training_inds]
  pred_train <- predictors[training_inds,]
  resp_test <- responses[-training_inds]
  pred_test <- predictors[-training_inds,]
  ols_mod <- lm(resp_train ~ as.matrix(pred_train))
  ridge_mod <- glmnet(as.matrix(pred_train), resp_train, alpha = 0, lambda = lambda)
  ols_in <- mean((resp_train - predict(ols_mod))^2)
  ols_out <- mean((resp_test - predict(ols_mod, newx=pred_test))^2)
  ridge_in <- mean((resp_train - predict(ridge_mod, newx=as.matrix(pred_train)))^2)
  ridge_out <- mean((resp_test - predict(ridge_mod, newx=as.matrix(pred_test)))^2)
  
  return(c(lambda = lambda, 
           ols_in = ols_in,     ols_out = ols_out, 
           ridge_in = ridge_in, ridge_out = ridge_out))
  
}
set.seed(1)
Shrink_results <- compare_ols_ridge(responses = GenCont[,1], 
                              predictors = GenCont[,-1], 
                              lambda=1)
```


## Exploring Lambda

Use your `compare_ols_ridge()` to look at the in- and out-of-sample prediction errors for a range of $\lambda$ from 0.1 to 100




```{r echo=FALSE, eval=params$show_answers, results="asis"}
cat("# Test statments")
```

```{r echo=params$show_answers, eval=params$show_answers}
scoreActivity::score253(14)
```

```{r echo = FALSE, eval = FALSE}
compare_ols_ridge(GenCont[,1], GenCont[,-1], lambda = 1)
```
