---
title: "In-Class Computing Task 15"
author: "Math 253: Statistical Computing & Machine Learning"
date: "Dimension Reduction Methods"
params:
  show_answers: FALSE
output: 
  rmarkdown::tufte_handout
---

```{r include=FALSE}
options(width=80)
library(mosaicData)
library(ggplot2)
library(scoreActivity)
library(ISLR)
library(pls)
library(glmnet)
knitr::opts_chunk$set(tidy=FALSE)
options(width=100)
```

In this activity, you're going to synthesize data from a linear system with multiple inputs and a single output, where only a few of the inputs contribute to the output.

## Overview

A linear system is often written

$${\mathbf Y} =  {\mathbf X} \cdot \beta + \epsilon$$

In this notation, $\beta$ are the *coefficients* of the system, ${\mathbf X}$ is the model matrix, and ${\mathbf Y}$ is the output.  ${\mathbf Y}$ is not completely set by ${\mathbf X} \cdot {\beta}$, there is some random part to ${\mathbf Y}$, unrelated to any of the columns in ${\mathbf X}$, represented by $\epsilon$.

We are going to make a seemingly strange choice of ${\mathbf X}$: the columns from a matrix representing a monochrome version of the Mona Lisa.^[The file is available at `"http://tiny.cc/dcf/mona.rda"` and contains a matrix `mona`.  I suggest you first download the file (`download.file()`) to your computer, with `destfile = "mona.rda"`. Do this just once and **don't** put the `download.file()` command in your `.R` script unless you have commented it out. It is meant to be run only once. In your script, you can load `mona` using `load("mona.rda")`.] This matrix has strong correlations among the columns.
```{r echo=params$show_answers}
load("../Class_notes/Images/mona.rda")
```

You will use the matrix `mona` to form ${\mathbf X}$ in your simulation.^[Why use the transpose operator, `t()`?  The row-column convention for images are reversed from those for matrices.  `t(mona)` gives a matrix with 250 rows and 191 columns.]
```{r createX}
X <- t(mona) - mean(mona[])
```
The `[]` in `mona[]` means to treat `mona` as one long vector rather than as a matrix.^[A matrix can be thought of as a collection of vectors.] 

\enlargethispage{1.5in}
Make two more matrices:

1. `X_rand` with has the same size as `X` but consists of iid N($0$,$1^2$) noise.
2. `X_corr` with columns that have the covariance as `X`.^[Remember, you can create correlated noise from iid noise by post-multiplying by the square-root of the desired covariance matrix.  That is, post-multiply `Xrand` with `chol(var(X))`.]

```{r echo=params$show_answers}
X_rand <- matrix(rnorm(191*250), ncol=191)
X_corr <- X_rand %*% chol(var(X))
```

```{r fig.margin=TRUE, fig.cap="Mona Lisa? Same data as the original, but the order of variables and cases has been randomized.", echo=FALSE}
image(X[sample(1:250, size=250),sample(1:191, size=191)], asp=1, xaxt='n', yaxt='n', bty="n")
```


## Sparse beta

There are 191 columns in each of `X`, `Xrand` and `Xcorr`.  Create a vector `beta` that has 191 numbers.  Of these 191 numbers, 175 should be 0.  The other 16 should have values of 2, 5, $-3$, or $-4$.  The order should be random. (Hints: Use these functions `rep()`, `sample(1:191)` as well as indexing.) A vector or matrix consisting mainly of zeros is called *sparse*.  The sparse $\beta$ you are using here simulates a system where there are many inter-related variables in ${\mathbf X}$, but just a few of them contribute to the formation of ${\mathbf Y}$

```{r echo = params$show_answers}
beta <- c(rep(0, 175), rep(2, 4), rep(5, 4), rep(-3, 4), rep(-4, 4))
beta <- sample(beta)
```

## The output

Create two output vectors based on ${\mathbf X} \cdot {\beta}$, using `X` for ${\mathbf X}$. Each of these output vectors will play the role of $Y$ in the linear system you are simulating.

\newpage

1. `Y_pure` which is simply ${\mathbf X} \cdot {\beta}$.
2. `Y_real` which is ${\mathbf X} \cdot {\beta}$ plus noise which is iid normal with mean 0 and a variance that is 10% of the variance of `Ypure`.

```{r echo = params$show_answers}
Y_pure <- X %*% beta
Y_real <- X %*% beta + rnorm(nrow(X), 0.1 * sd(X[]))
```

## Least squares

Use `lm()` to fit `Y_pure` against `X`. Create a vector `beta_hat_pure` to hold the coefficients.  Plot `beta_hat_pure` against `beta` and draw a conclusion about the performance of `lm()` in this case.^[Remember that your `X` has no vector corresponding to the intercept.  Discard the intercept from the coefficients found by `lm()`.]

```{r echo=params$show_answers, eval=params$show_answers, fig.margin=TRUE,fig.cap="\\texttt{beta\\_hat\\_pure} is an almost exact match to \\texttt{beta}."}
mod_lm <- lm(Y_pure ~ X)
beta_hat_pure <- coef(mod_lm)
plot(beta, beta_hat_pure[-1])
```

Now do the same for `Y_real`, creating a vector of fitted coefficients `beta_hat_real` and comparing that to your actual `beta`.

```{r echo=params$show_answers, eval=params$show_answers, fig.margin=TRUE,fig.cap="\\texttt{beta\\_hat\\_real} has little relationship to $\\beta$."}
mod_lm_real <- lm(Y_real ~ X)
beta_hat_real <- coef(mod_lm)
plot(beta, beta_hat_real[-1])
```

Would you be able to detect from `beta_hat_real` that ${\beta}$ is sparse?

## The lasso estimator

Use the lasso method to estimate ${\hat{\beta}}$, which you can store in a vector `beta_lasso`. The `glmnet` package has a command `cv.glmnet()` which uses cross-validation to choose an appropriate value of $\lambda$. The commands look like this:

```{r cache=TRUE}
library(glmnet)
lasso_mod <- cv.glmnet(X, Y_real, alpha=1)
beta_lasso <- predict(lasso_mod, type = "coefficients", s = lasso_mod$lambda.min)
```


```{r echo=params$show_answers, eval=params$show_answers, fig.margin=TRUE}
plot(jitter(beta), beta_lasso[-1] ) # [-1] get rid of intercept
```

\enlargethispage{1.5in}

## Principal components

Recall that each of the principal components of a matrix has a scalar --- called the *singular value* --- indicating the "size" of that principal component in contributing to the reconstruction of the matrix.  You can find these scalars like this:
```{r}
sing_vals <- svd(X)$d
```

```{r eval = TRUE, fig.margin=TRUE, fig.cap="Linear Lisa? This is \\texttt{X\\_corr} and has the same covariance as \\texttt{mona}.", echo=FALSE}
image(X_corr, asp=1, xaxt='n', yaxt='n', bty="n")
```

```{r fig.margin=TRUE, fig.cap="Iid Lisa?", echo=FALSE}
image(X_rand, asp=1, xaxt='n', yaxt='n', bty="n")
```


The cumulative sum `cumsum(sing_vals^2)` divided by `sum(sing_vals^2)` produces the $R^2$ of the approximation of `X` using successively $k = 1, 2, 3, ..., 191$ principal components.

Find the singular values of `X_rand` and `X_corr`. On the same graph, plot out $R^2$ versus $k$ for the singular values from each of  `X`, `X_rand`, and `X_corr`.

Calculate how many principal components are needed to reconstruct the matrix with an $R^2$ of 99%.  Call your answers `n99`, `n99_rand` and `n99_corr` respectively.

```{r echo = params$show_answers}
sing_vals_rand <- svd(X_rand)$d
sing_vals_corr <- svd(X_corr)$d
Rsq <- cumsum(sing_vals^2) / sum(sing_vals^2)
Rsq_rand <- cumsum(sing_vals_rand^2) / sum(sing_vals_rand^2)
Rsq_corr <- cumsum(sing_vals_corr^2) / sum(sing_vals_corr^2)
n99 <- 1 + sum(Rsq < 0.99)
n99_rand <- 1 + sum(Rsq_rand < 0.99)
n99_corr <- 1 + sum(Rsq_corr < 0.99)
```



\newthought{Finally, use principal components to model} `Y_real` against `X`.  The commands look like this.

```{r}
library(pls)
pcr.fit <- pcr(Y_real ~ X, scale = TRUE, validation="CV") 
```

Using `R2(pcr.fit)`, examine the cross-validated $R^2$ as a function of the number of components used.  How many components are needed to get to, say, $R^2 = 0.85$.  


```{r echo=FALSE, eval=params$show_answers, results="asis"}
cat("# Test statments")
```

```{r echo=params$show_answers, eval=params$show_answers}
source("Day-15-Test-Statements.R", local = TRUE)
```
