---
title: "In-Class Programming Task 10"
author: 'Math 253: Statistical Computing & Machine Learning'
date: "Implementing logistic regression"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    latex_engine: xelatex
  tufte::tufte_book:
    latex_engine: xelatex
params:
  show_answers: no
---

```{r include=FALSE}
options(width=80)
library(mosaicData)
library(ggplot2)
library(scoreActivity)
library(ISLR)
knitr::opts_chunk$set(tidy=FALSE, echo=params$show_answers)
options(width=100)
```

In this activity, you will work with the logistic function and likelihood. For the sake of definiteness, you'll use the `Default` data from the `ISLR` package, but the programming involved will show techniques for writing functions that work with any data table.

```{r}
library(ISLR)
data(Default)
```

# The logistic function

The logistic function is defined as ${\cal L}(x) = \frac{e^x}{1 + e^x}$. Note that the function ${\cal L}(x)$ takes a scalar $x$ as an input.

* Write a function `logistic()` that returns the value of the logistic function for any numerical vector `x`. Remember that functions like `exp()` and `+` are vectorized in R, so writing `logistic()` for a scalar $x$ will automatically work on a vector `x`.

```{r echo = params$show_answers}
logistic <- function(x) exp(x) / (1 + exp(x))
```



# Linear combinations

In logistic regression, a linear combination of variables is taken and passed through the logistic function to produce a probability. 

* Write a function `linear_combine()` that takes two arguments:    
    - `data=` specifying a data table
    - `coefs=` specifying a set of coefficients to use in the linear combination.

The `coefs` argument will have a very specific form: a numerical vector with named components.  For example, `linear_combine()` will be invoked with a command like this:
```{r eval=FALSE, echo=TRUE}
linear_combine(data = Default, 
               coefs = c(intercept = 3, balance = 0.00001, income = -0.0002))
```

The `linear_combine()` function will multiply each of the named variables in `data` by the corresponding coefficient, then add in the value of the intercept.  The returned value will be the vector that is the specified linear combination.

\enlargethispage{1in}
At the heart of your `linear_combine()` will be a loop like this:
```{r eval=FALSE, echo=TRUE}
result <- 0
for (nm in names(coefs)) {
  if (nm == "intercept") {
    result <- result + coefs[[nm]]
  } else {
    result <- result + coefs[[nm]] * data[[nm]]
  }
}
```
    
Arrange your function to call `stop()` if the variable names in the `coef` vector don't exist in `data`.

```{r echo=params$show_answers}
linear_combine <- function(data = NULL, coefs = NULL) {
  if (is.null(data)) 
    stop("Must provide data table in data= argument")
  if (is.null(coefs)) 
    stop("Must provide coefficients in coef= argument")
  result <- 0
  for (nm in names(coefs)) {
    if (nm == "intercept") {
      result <- result + coefs[[nm]] + rep(0, nrow(data))
    } else {
      if ( ! nm %in% names(data)) 
        stop("Variable '", nm, "' in coefficients but not in data.")
      result <- result + coefs[[nm]] * data[[nm]]
    }
  }
  result
}
```

# Probabilities

\enlargethispage{1in}

Applying the logistic function to the output of `linear_combine()` produces a number between zero and one.  You can interpret this probability as the conditional probability of the outcome given the values of the data. 

* Write a function `LL_logistic()` that computes the log-likelihood of the observed result given a model in the form of `coefs`.  It will have these arguments:
    * `data=` a data table for training
    * `coefs=` a proposed vector of coefficients
    * `outcome=` a `TRUE` or `FALSE` vector giving the observed outcome in the training data.
    
Your function should 

1. use `linear_combine()` on `data` and `coefs` to produce a number for each case in `data`.
2. put the result of (1) through `logistic()`.  The output will be a probability for each case in `data`, which I'll write here as $p_i$.
3. set the likelihood for each case in `data` to be $p_i$ if the result is `TRUE`, otherwise set the likelihood to be $1 - p_i$.
4. combine the likelihoods of the individual cases into an overall log likelihood.

```{r echo=params$show_answers}
LL_logistic <- function(data = NULL, coefs = NULL, outcome) {
  probs <- logistic( linear_combine(data=data, coefs=coefs))
  likelihoods <- ifelse(outcome, probs, 1-probs)
  
  sum(log(likelihoods))
}
```

Here's a quick test for your function:

```{r echo=TRUE}
LL_logistic(data=Default, 
            coefs = c(intercept = 1, income = -0.0001), 
            outcome = Default$default == "Yes")
```

# Optimize

Use `optim()` to find the maximum likelihood coefficients (using just the intercept and `income`) to predict `default`. 

```{r echo=params$show_answers, eval=params$show_answers}
fun_to_minimize <- function(coefs) {
  - LL_logistic(data=Default, coefs=coefs, outcome=Default$default=="Yes")
}
best_coefs <- optim(c(intercept = 1, income = -0.0001), fun_to_minimize) 
```

Put your results in an object called `best_coefs`.

Compare your result to the output of 
```{r eval=FALSE, echo=TRUE}
glm(default == "Yes" ~ income, data=Default, family="binomial")
```

```{r echo=FALSE, eval=params$show_answers, results="asis"}
cat("# Test statments")
```

```{r echo=params$show_answers, eval=params$show_answers}
scoreActivity::score253(10)
```


\newpage

# Above and beyond

It seems odd to have to specify both `data=Default` and `outcome=Default$default == "yes"` in the above command.  R provides a way to avoid this, so that the `outcome` argument can be simply `default == "Yes"`.  To do this, write your `LL_logistic()` this way

```{r eval=FALSE, echo=TRUE}
LL_Logistic <- function(data = NULL, coefs = NULL, outcome) {
  outcome_statement <- substitute(outcome)
  outcome <- eval(outcome_statement, envir = data)
  # Now, "outcome" will be the value of the statement when 
  # applied to the values in data
  
  # ... Continue with your calculations
}
```

