---
title: "In-class Programming Activity 12"
author: "Math 253: Statistical Computing & Machine Learning"
date: "Testing LDA and QDA"
params: 
  show_answers: FALSE
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    latex_engine: xelatex
  tufte::tufte_book:
    latex_engine: xelatex
---

```{r include=FALSE}
options(width=80)
library(mosaicData)
library(ggplot2)
library(scoreActivity)
library(ISLR)
library(scales) # for simple transparency in colors
knitr::opts_chunk$set(tidy=FALSE, echo=params$show_answers)
options(width=100)
```

Today, you're going to generate simulated data to test out LDA and QDA. The data set you make will have two quantitative variables `x` and `y` to use as inputs, and a output `color` with levels `"red"` and `"blue"`. Actually, you'll be creating two different data tables:

* `Same_sigma` where cases in the two classes have the same $(x,y)$ covariance matrix for the predictor variables.
* `Different_sigma` where the two classes have different $(x,y)$ covariance matrices.

# Generating simulated data


* Create an object `n_cases` with the value 100.
* Create an object `red_mean` with the value `c(1, 0)`.
* Create an object `green_mean` with value `c(0, -1)`.
* Create an object `blue_mean` with value `c(-1, 1)`.
* Create an object `covar_red` which is a 2 x 2 matrix with 3 and 1 on the diagonal and $-1.7$ on the off-diagonal. It should be structured as a covariance matrix.
* Similarly, create an object `covar_blue`, a covariance matrix with 2 and 3 on the diagonal and $1.5$ on the off-diagonal.
* Create three matrices, `One`, `Two`, and `Three`.  Each should have the appropriate shape for data from `n_cases` cases and two variables. The variables should each be `n_cases` random draws from a $\mbox{N}(0, 1^2)$ distribution, that is, a normal distribution with mean zero and standard deviation $1$.^[In specifying the normal distribution, one needs to decide whether to report the standard deviation or the variance.  R uses `sd=`.  To help to eliminate ambiguity in mathematical notation, the form $1^2$ is used simply as a reminder that the quantity is a variance.]
* Now you're going to transform `One`, `Two` and `Three` to create correlations between the two columns.     
    a -- Create an matrix `Red` that is `One` times the Cholesky decomposition of `covar_red`.  The `Red` matrix will contain correlated random variables with a covariance of approximately `covar_red`.
    
    b -- Similarly, create a matrix `Green` which will be `Two` times the Cholesky decomposition of `covar_red`. That is, `Green` and `Red` will have the same correlation structure.

    c -- Now create a matrix `Blue` which will be `Three` times the Cholesky decomposition of the other covariance matrix, `covar_blue`.

\enlargethispage{1in}

* Modify the `Red`, `Green` and `Blue` matrices by adding to each column a value for the mean drawn from `red_mean`, `green_mean`, and `blue_mean` respectively.^[*Hint*: It won't work to do the obvious, simple thing, e.g.  add `red_mean` to `Red`. There are many ways to construct a statement that works.  Among others, there's a way using `outer()`, a way using `matrix()`, and even a way using `t()` twice.] That is, for `Red`, add 1 to the first column and 0 to the second.

\newpage

* Create three data frames, each with variables `x`, `y`, and `class`^[You'll use `data.frame()` to construct the data frames.  Make sure to give the optional argument `stringsAsFactors = FALSE`.  This will let the class be stored as straightforward character strings that can be used in plotting to specify the color.]
    
    a -- `Red` will have `x` as the first column of `Red`, `y` as the second column of `Red`, and `class` set equal to the string `"red"`.
    
    b --  `Blue` will be the same thing but using the columns of `blue` and the `class` set to `"blue"`.
    
    c --  `Green` is similar, using the columns of `Green` and the class `"green"`.
* Last step in generating the simulated data: make two data frames each of which combines "data" from two classes.
```{r echo=TRUE, eval=FALSE}
Same_sigma <- rbind(Red, Green)
Different_sigma <- rbind(Red, Blue)
```

```{r echo=params$show_answers}
n_cases <- 10000
red_mean <- c(1, 0)
green_mean <- c(0, -1)
blue_mean <- c(-1, 1)
covar_red <- matrix(c(3, -1.7, -1.7, 2), nrow=2, ncol=2)
covar_blue <- matrix(c(2, 1.5, 1.5, 3), nrow=2, ncol=2)
One <- matrix(rnorm(2 * n_cases), ncol=2, nrow=n_cases)
# other ways to do this
#       cbind(rnorm(n_cases), rnorm(n_cases))
#       matrix(rnorm(2 * n_cases), ncol=2)
Two <- matrix(rnorm(2 * n_cases), ncol=2)
Three <- matrix(rnorm(2 * n_cases), ncol=2, nrow=n_cases)
Red <- One %*% chol(covar_red) + outer(rep(1, n_cases), red_mean)
# Another way to generate the matrix to the right of +
#      matrix(red_mean, nrow = n_cases, ncol = 2, byrow = TRUE)
Green <- Two %*% chol(covar_red) + outer(rep(1, n_cases), green_mean)
Blue <- Three %*% chol(covar_blue) + outer(rep(1, n_cases), blue_mean)
Red <- data.frame(x = Red[,1], y = Red[,2], class="red", 
                  stringsAsFactors = FALSE)
Green <- data.frame(x = Green[,1], y = Green[,2], class="green",
                    stringsAsFactors = FALSE)
Blue <- data.frame(x = Blue[,1], y = Blue[,2], class="blue",
                   stringsAsFactors = FALSE)
Same_sigma <- rbind(Red, Green)
Different_sigma <- rbind(Red, Blue)
```
```{r echo=params$show_answers, fig.margin=TRUE, eval=params$show_answers, fig.height=5, fig.width=5,out.width="2.5in"}
plot(y ~ x, data=Same_sigma, col=alpha(Same_sigma$class, .02), pch=20)
```
```{r echo=params$show_answers, fig.margin=TRUE, , fig.height=5, fig.width=5,out.width="2.5in", fig.cap="The two classes of cases, red and blue, in \\texttt{Sim\\_two}."}
plot(y ~ x, data=Different_sigma, col=alpha(Different_sigma$class, .02), pch=20)
```

# LDA and QDA

Fit a linear discriminant model `class ~ x + y` to the data in `Same_sigma`. Call the model `mod_LDA_one`.
```{r echo=TRUE}
library(MASS)
mod_LDA_one <- lda(class ~ x + y, data = Same_sigma)
```

Then use the model to test the model on the same training data to which it was fit. Store the result in `test_LDA_one`.
```{r echo=TRUE}
test_LDA_one <- predict(mod_LDA_one, newdata = Same_sigma)
```

The resulting object, `test_LDA_one`,  is a list of three items. Make sure you understand what each of them is.

QDA works in the same way: the function is `qda()`.

# Confusion matrices

The confusion matrix compares the actual class to the predicted class from the model.  It's straightforward to compute:
```{r echo=TRUE, eval=params$show_answers}
table(Same_sigma$class, test_LDA_one$class)
```

* Compare the confusion matrix from LDA on `Same_sigma` to that from QDA based on `Same_sigma`.^[Use the names `mod_QDA_one` and `test_QDA_one` to store the fitted model and the test results from `predict()` respectively.]  Which one shows better performance?
```{r echo=params$show_answers, eval=params$show_answers}
mod_QDA_one <- MASS::qda(class ~ x + y, data = Same_sigma)
test_QDA_one <- predict(mod_QDA_one, newdata = Same_sigma)
table(Same_sigma$class, test_QDA_one$class)
```

* Fit both LDA and QDA models to `Sim_two`.  Which one performs better?
```{r echo=params$show_answers, eval=params$show_answers}
mod_LDA_two <- MASS::lda(class ~ x + y, data = Different_sigma)
mod_QDA_two <- MASS::qda(class ~ x + y, data = Different_sigma)
test_LDA_two <- predict(mod_LDA_two, newdata = Different_sigma)
test_QDA_two <- predict(mod_QDA_two, newdata = Different_sigma)
table(Different_sigma$class, test_LDA_two$class)
table(Different_sigma$class, test_QDA_two$class)
```

# Bigger n

The difference in performance of LDA and QDA in these examples is not so large that it's evident in a sample with 100 cases of each class.  Go back and set `n_cases` to be 10000, and re-evaluate the confusion matrices.

\enlargethispage{1in}

# Above and beyond

Calculate the log likelihood for `mod_LDA_one` against the observations `Different_sigma$class`.

```{r echo=params$show_answers, eval=params$show_answers}
likelihoods <-
  with(test_LDA_one, 
       ifelse(class == "red", posterior[,"red"], posterior[,"green"]))
sum(log(likelihoods))
```


```{r echo=FALSE, eval=params$show_answers, results="asis"}
cat("# Test statments")
```

```{r echo=params$show_answers, eval=params$show_answers}
scoreActivity::score253(12)
```
