# Linear Regression {#linear-regression}


```{r include=FALSE}
require(mosaic)
```

## Day 4 Preview

* The linear model (e.g. what `lm()` does)
* A variety of questions relevant to different purposes, e.g.
    - how good will a prediction be?
    - what's the strength of an effect?
    - is there synergy between different factors?


### ISL book's statement on why to study linear regression

> "Though it may seem somewhat dull compared to some of the more modern statistical learning approaches described ... later ..., linear regression is still a useful and widely used statistical learning method.  Moreover, it serves as a good jumping-off point for newer approaches....  Consequently, the importance of having a good understanding of linear regression before studying more complex learning methods cannot be overstated."

Concepts from linear regression:

* Compact representation of model form: polynomial coefficients.
* Much of inference (confidence intervals, hypothesis tests) can be expressed in terms of a polynomial coefficient.
* "Size" of model quantifiable as an integer: number of coefficients: degrees of freedom.
* Highly efficient estimation (when doing least squares)


## Small data

The regression techniques were developed in an era of small data, such as that that might be written in a lab notebook or field journal.  As a result:

1. Emphasis on very simple descriptions, such as means, differences between means, simple regression.
2. Theoretical concern with details of distributions, such as the t-distribution. 
    - the difference between z- and t-distributions are of no consequence for moderate DF and higher.
3. No division into training and testing data.  Data are too valuable to test! (Ironic, given the importance of replicability in the theory of the scientific method.)

As a consequence of (3), there's a great deal of concern about *assumptions*, e.g. 

* linearity of $f({\mathbf X})$
* structure of $\epsilon$: IID --- Independent and Identically Distributed
    - uncorrelated between cases
    - each is a draw from the same distribution.



## Programming basics: Linear Models

Syntactic element: **formulas**. Formulas provide a way of using variables "symbolically." This is useful, for instance, in depicting the desired relationship among variables. Two forms:

- `y ~ x` two-sided
- `~ x` one-sided
- NOT ALLOWED, `y ~`

Important functions: 

- `lm()`, `predict()` , `anova()`, `summary()`.
- For later: `solve()`, `model.matrix()`.
- From 155: `coef()`, `fitted()`, `resid()`

Training with `lm()`: Specify formula `Y ~ X1 + X2` ...

```{r}
data(College, package = "ISLR")
mod <- lm(Outstate ~ Enroll + Accept + perc.alumni, data = College ) 
coef(mod)
```

What kind of thing is `mod`?

Model output with `predict()`
```{r}
predict(mod, 
        newdata = data.frame(Enroll = 100, Accept = 1000, perc.alumni = 25))
```

What kind of thing is the output of `predict()`?

```{r}
predict(mod, interval = "confidence",
        newdata = data.frame(Enroll = 100, Accept = 1000, perc.alumni = 25))
predict(mod, interval = "prediction",
        newdata = data.frame(Enroll = 100, Accept = 1000, perc.alumni = 25))
```

Why is the "confidence interval" so much narrower than the "prediction interval?"

**Inference** with `anova()` and `summary()`. This is all "in-sample" inference, not cross-validated.

```{r}
M <- model.matrix(~ Enroll + Accept * perc.alumni, data = College)
qr.solve(M, College$Outstate)
```

For the people who have had linear algebra, why doesn't this work?
```{r error = TRUE}
solve(M, College$Outstate)
```


[Indexing on data: training and testing data sets](../ProgrammingActivities/Day-03-Programming.html)


## Review of Day 4, Sept 15, 2016

* Discussed the linear regression architecture and how it relates to machine learning.
    - linear regression is designed to work even in "small data" situations where
        a. cross-validation is not appropriate
        b. the number of model degrees of freedom may be almost as large as $n$
    - provides a ready definition of the "size" of a model: the number of coefficients.
* Introduced the main software used in linear regression, `lm()` and `predict()`
* Programming basics: indexing of vectors, matrices and data frames.

## Selecting model terms

The regression techniques 

- Heirarchical principal
    - main effects, then
    - interaction.
- Increase in $R^2$


### Theory of whole-model ANOVA.

Standard measure: $\frac{\mbox{Explained amount}}{\mbox{Unexplained amount}}$

Examples:

- Standard error of mean:  $\frac{\hat{\mu}}{\sigma / n}$ -- note the $n$.
- t statistic on difference between two means: $\frac{\hat{\mu}_1 - \hat{\mu}_2}{\sigma / (n-1)}$
- F statistic: $\frac{SS / df1}{SSR / df2}$
    - df1 is the number of degrees of freedom involved by the model or model term under consideration.
    - df2 is $n - (p - 1)$ where $p$ is the total degrees of freedom in the model.  (I called this $m$ in the Math 155 book.)  The intercept is what the $-1$ is about: the intercept *can never* account for case-to-case variation.

Trade-off between eating variance and consuming degrees of freedom.

The $R^2$ versus $p$ picture.

- Adjusted $R^2$
- Whole model ANOVA.
- ANOVA on model parts

## Programming basics: Graphics

Basic functions:

1. Create a frame: `plot()`.  Blank frame: `plot( , type="n")`
    - set axis limits, 
2. Dots: `points(x, y)`, `pch=20`
3. Lines: `lines(x, y)` --- with `NA` for line breaks
4. Polygons: `polygon(x, y)` --- like lines but connects first to last.
    - fill
5. Color, size, ... `rgb(r, g, b, alpha)`, "tomato"



## In-class programming activity 

[Programming Activity 4](../Daily-Programming/Day-4-Programming-Task.pdf): Drawing a histogram.


## Review of Day 4

### Graphics basics

1. API for graphics: `plot()`, `points()`, `lines()`, `polygon()`, `text()`, ...
2. Create a plotting frame: `plot()`
    - Write a function that makes this more convenient to use. What features would you like. 
    ```{r}
    blank_frame <- function(xlim, ylim) {
      
    }
    ```
3. Write a function to draw a circle.
    - What do you want the interface to look like? What arguments are essential? What options are nice to have?
    
## Regression and Interpretability

Regression models are generally constructed for the sake of interpretability:

* Global linearity
* Coefficients are indication of effect size. The coefficients have physical units.
* Term by term indication of statistical significance

## Measuring Accuracy of the Model

* $R^2$ - Var(fitted)/Var(response)
* Adjusted $R^2$ - takes into account estimate of average increase in $R^2$ per junk degree of freedom
* Residual Standard Error - Sqrt of Average square error per residual degree of freedom. The sqrt of the mean square for residuals in ANOVA

## Bias of the model

* Perhaps effect of TV goes as sqrt(money) as media get saturated?
* Perhaps there is a synergy that wasn't included in the model?


- Whole model ANOVA.
- ANOVA on model parts
- Adjusted $R^2$

Run an example on `College` data from `ISLR` package 

```{r}
data(College, package="ISLR")
College$Yield <- with(College, Enroll/Accept)
mod1 <- lm(Yield ~ Outstate + Grad.Rate + Top25perc, data=College)
```
* What variables matter?
* How good are the predictions?
* How strong are the effects?



## Forward, backward and mixed selection

Use the `College` model to demonstrate each of the approaches by hand.  Start with `pairs()` or write an `lapply()` for the correlation with `Yield`?

Create a whole bunch of model terms

- "main" effects
- "interaction" effects
- nonlinear transformations: powers, logs, sqrt, steps, ...
- categorical variables

Result: a set of $k$ vectors that we're interested to use in our model.

Considerations:

- not all of the $k$ vectors may pull their weight
- two or more vectors may overlap in how they eat up variance

Algorithmic approaches:

- Try all combinations, pick the best one.
    - computationally expensive/impossible $2^k$ possibilities
    - what's the sensitivity of the process to the choice of training data?
- "Greedy" approaches


## In-class programming activity 

[Day 5 activity](../Daily-Programming/Day-05-Programming-Task.pdf)

Drawing a histogram.


