---
title: "Topic 4: Classifiers"
author: 'Math 253: Statistical Computing & Machine Learning'
date: "Daniel Kaplan"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    latex_engine: xelatex
  tufte::tufte_book:
    latex_engine: xelatex
---

## Classification overview

* Response variable: categorical.  Typically just a few levels: 2 or 3.
* Two types of outputs from classification models:     
    1. The predicted category given the inputs
    2. Probability of each category given the inputs
        - Type (2) can be fitted with maximum likelihood.
* Trade-offs:     
    * Flexibility vs interpretability
    * Accuracy vs bias
* Four model architectures
    1. Logistic regression.  Especially important for interpretability.
    2. Linear discriminant analysis
    3. Quadratic discriminant analysis
    4. K nearest neighbors

## Day 10 preview

1. Probability and odds    
    * [Theme Song](https://www.youtube.com/watch?v=RthEYvh6aMM)
    * [Making book](http://en.wikipedia.org/wiki/Dutch_book)
2. Multivariate gaussians (Maybe)
3. Programming activity: Poker hands

# Logistic regression

```{r child = "411-Logistic-Regression.Rmd"}
```

```{r child = "412-LDA-QDA.Rmd"}
```

